---
title: "p8105_hw2_cl4776"
author: "Chenxi Liu"
date: "2025-09-30"
output: github_document
---

```{r}
library(tidyverse)
library(readxl)
library(lubridate)
library(janitor)
```

# Problem 1

## clean the data in pols_month.csv

```{r}
pols = read_csv("data/pols-month.csv") |> 
   separate(mon, into = c("year", "month", "day"), sep = "-") |>
   mutate(
    year = as.integer(year),
    month = as.integer(month),
    day = as.integer(day),
   
    month = month.name[month],
  
    president = case_when(
      prez_gop == 1 ~ "gop",
      prez_dem == 1 ~ "dem"
    )
  ) |>
  select(-day, -prez_gop, -prez_dem)
pols
```

## clean the data in snp.csv

```{r}
snp = read_csv("data/snp.csv") |>
  mutate(
    date  = mdy(date),
    year  = year(date),
    month_num = month(date),
    month = factor(month.name[month_num], levels = month.name) 
  ) |>
  arrange(year, month_num) |>
  select(year, month, close, -date, -month_num)
snp
```

## Tidy the unemplyment data

```{r}

unemployment = read_csv("data/unemployment.csv") |>
  pivot_longer(
    Jan:Dec, 
    names_to = "month", 
    values_to = "unemployment"
  ) |>
  mutate(
    year = as.integer(Year),
    month = month.name[match(month, month.abb)]
  ) |>
  select(year, month, unemployment) |>
  arrange(year, match(month, month.name))
unemployment
```

## Merge with previous data and describle these three datasets

```{r}
pols_snp = left_join(pols, snp, by = c("year", "month"))

final_data = left_join(pols_snp, unemployment, by = c("year", "month"))

dim(final_data)      
range(final_data$year)  
names(final_data)    
```
The **pols-month** dataset contains counts of national politicians by party and the president’s affiliation, the **snp** dataset provides monthly closing values of the S&P 500 index, and the **unemployment** dataset gives monthly unemployment rates. After tidying and merging by year and month, the resulting dataset has `r nrow(final_data)` rows and `r ncol(final_data)` columns, covers the years `r min(final_data$year)`–`r max(final_data$year)`, and includes key variables such as `year`, `month`, `president`, `close`, and `unemployment`.

# Problem 2


## Read and clean Mr. Trash Wheel data

```{r}

mr_trash = read_excel(
    "data/202409 Trash Wheel Collection Data.xlsx",
    sheet = "Mr. Trash Wheel",
    skip = 1        
  ) |>
  clean_names() |>
  filter(!is.na(dumpster)) |>               
  mutate(
    sports_balls = as.integer(round(sports_balls)),
    trash_wheel = "Mr"
  )
mr_trash
```


