---
title: "p8105_hw2_cl4776"
author: "Chenxi Liu"
date: "2025-09-30"
output: github_document
---

```{r}
library(tidyverse)
library(readxl)
library(lubridate)
library(janitor)
```

# Problem 1

## clean the data in pols_month.csv

```{r}
pols = read_csv("data/pols-month.csv") |> 
   separate(mon, into = c("year", "month", "day"), sep = "-") |>
   mutate(
    year = as.integer(year),
    month = as.integer(month),
    day = as.integer(day),
   
    month = month.name[month],
  
    president = case_when(
      prez_gop == 1 ~ "gop",
      prez_dem == 1 ~ "dem"
    )
  ) |>
  select(-day, -prez_gop, -prez_dem)
pols
```

## clean the data in snp.csv

```{r}
snp = read_csv("data/snp.csv") |>
  mutate(
    date  = mdy(date),
    year  = year(date),
    month_num = month(date),
    month = factor(month.name[month_num], levels = month.name) 
  ) |>
  arrange(year, month_num) |>
  select(year, month, close, -date, -month_num)
snp
```

## Tidy the unemplyment data

```{r}

unemployment = read_csv("data/unemployment.csv") |>
  pivot_longer(
    Jan:Dec, 
    names_to = "month", 
    values_to = "unemployment"
  ) |>
  mutate(
    year = as.integer(Year),
    month = month.name[match(month, month.abb)]
  ) |>
  select(year, month, unemployment) |>
  arrange(year, match(month, month.name))
unemployment
```

## Merge with previous data and describle these three datasets

```{r}
pols_snp = left_join(pols, snp, by = c("year", "month"))

final_data = left_join(pols_snp, unemployment, by = c("year", "month"))

dim(final_data)      
range(final_data$year)  
names(final_data)    
```
The **pols-month** dataset contains counts of national politicians by party and the president’s affiliation, the **snp** dataset provides monthly closing values of the S&P 500 index, and the **unemployment** dataset gives monthly unemployment rates. After tidying and merging by year and month, the resulting dataset has `r nrow(final_data)` rows and `r ncol(final_data)` columns, covers the years `r min(final_data$year)`–`r max(final_data$year)`, and includes key variables such as `year`, `month`, `president`, `close`, and `unemployment`.

# Problem 2


## Read and clean Mr. Trash Wheel data

```{r}


mr_trash  = read_excel(
    "data/202409 Trash Wheel Collection Data.xlsx",
    sheet = "Mr. Trash Wheel",
    skip = 1        
  ) |>
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |>               
  mutate(
    sports_balls = as.integer(round(sports_balls)),
    trash_wheel = "Mr",
    year         = as.integer(year)    
  )
```
## Professor Trash Wheel

```{r}
prof_trash = read_excel(
    "data/202409 Trash Wheel Collection Data.xlsx",
    sheet = "Professor Trash Wheel",
    skip = 1
  ) |>
  janitor::clean_names() |>
  filter(!is.na(dumpster))

if(!"sports_balls" %in%  names(prof_trash)) {
  prof_trash = prof_trash |> mutate(sports_balls = NA_integer_)
}

prof_trash = prof_trash |> mutate(
  trash_wheel = "Professor",
  year        = as.integer(year)
)
                            

```

## Gwynnda Trash Wheel

```{r}
gwynnda_trash = read_excel(
    "data/202409 Trash Wheel Collection Data.xlsx",
    sheet = "Gwynnda Trash Wheel",
    skip = 1
  ) |>
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |>
  select(where(~!all(is.na(.)))) 

if(!"sports_balls" %in% names(gwynnda_trash)) {
  gwynnda_trash = gwynnda_trash |> mutate(sports_balls = NA_integer_)
}

gwynnda_trash = gwynnda_trash |> mutate(
  trash_wheel = "Gwynnda",
   year        = as.integer(year) 
  )

```

## Combine these three datasets

```{r}
trash_wheel = bind_rows(mr_trash, prof_trash, gwynnda_trash)
trash_wheel %>%
  dplyr::glimpse()
```

The combined Trash Wheel dataset contains information from Mr. Trash Wheel, Professor Trash Wheel, and Gwynnda, with `r nrow(trash_wheel)` observations and `r ncol(trash_wheel)` variables. Key variables include the dumpster identifier (`dumpster`), collection date (`year`, `month`), total weight of trash collected in tons (`weight_tons`), and item-specific counts such as `cigarette_butts`, `plastic_bottles`, and `sports_balls`. In total, Professor Trash Wheel has collected `r sum(trash_wheel$weight_tons[trash_wheel$trash_wheel == "Professor"], na.rm = TRUE)` tons of trash, while Gwynnda collected `r sum(trash_wheel$cigarette_butts[trash_wheel$trash_wheel == "Gwynnda" & trash_wheel$year == 2022 & trash_wheel$month == "June"], na.rm = TRUE)` cigarette butts in June 2022.

# Problem 3

```{r}
library(tidyverse)
library(janitor)
library(lubridate)
```

## ---------- 1. Zillow ZORI NYC data ----------
```{r}
zori = read_csv("data/zillow_data/Zip_zori_uc_sfrcondomfr_sm_month_NYC.csv") |>
  clean_names() |>
  rename(
    zip = region_name,
    county = county_name
  ) |>
  mutate(zip = str_pad(as.character(zip), 5, pad = "0")) |>
  pivot_longer(
    cols = matches("^x\\d{4}_\\d{2}_\\d{2}$"),   
    names_to = "date",
    values_to = "zori"
  ) |>
  mutate(
    date  = ymd(str_remove(date, "^x")),  
    year  = year(date),
    month = month(date, label = TRUE, abbr = FALSE)
  ) |>
  select(zip, county, city, state, date, year, month, zori)

```

## ---------- 2. NYC ZIP code ↔ Borough/Neighborhood ----------
```{r}

nyc_map = read_csv("data/zillow_data/Zip Codes.csv") |>
  clean_names() |>
  rename(
    zip = zip_code,         
    borough = county,       
    neighborhood = neighborhood
  ) |>
  mutate(zip = str_pad(as.character(zip), 5, pad = "0")) |>
  select(zip, borough, neighborhood)

```

## ---------- 3. Merge datasets ----------
```{r}
final <- zori |>
  inner_join(nyc_map, by = "zip") |>
  relocate(zip, borough, neighborhood, county, date, year, month, zori)
```
## ---------- 4. Quick summaries ----------
```{r}
n_obs  <- nrow(final)
n_zip  <- n_distinct(final$zip)
n_hood <- n_distinct(final$neighborhood)

missing_zips <- nyc_map |> anti_join(zori, by = "zip")
```
## Dataset description

```{r}
n_obs  <- nrow(final)
n_zip  <- n_distinct(final$zip)
n_hood <- n_distinct(final$neighborhood)

n_obs; n_zip; n_hood
```
After cleaning and merging Zillow’s ZORI data with the NYC ZIP–borough–neighborhood crosswalk, the final tidy dataset contains `r n_obs` observations across `r n_zip` unique ZIP codes and `r n_hood` unique neighborhoods.  
Key variables include `zip`, `borough`, `neighborhood`, `date` (with `year` and `month`), and `zori` (the Zillow Observed Rent Index).


